---
title: "Fusion Gene Detection Pipeline"
subtitle: "Structural Variant Analysis and Gene Fusion Prediction"
date: "`r Sys.Date()`"
author: "Bellerophon Pipeline"
format: 
  html:
    toc: true
    toc-depth: 3
    self-contained: true
    code-fold: true
    code-summary: "Show code"
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
params:
  input_file: "data/example_input/sample_variants.mavis.tsv"
  input_format: "auto"  # "auto", "vcf", "mavis", "hgvs"
  output_dir: "output"
  genome_build: "GRCh38"
  gencode_release: "49"
  disease_gene_list: "data/annotations/Nijmegen.DG.ENSG.list.txt"
  gtex_file: "data/annotations/GTEx_Analysis_2025-08-22_v11_RNASeQCv2.4.3_gene_median_tpm.gct.gz"
  ncbi_contig_mapping: "data/ncbi_contig_mapping.tsv"
editor_options: 
  chunk_output_type: inline
editor: 
  markdown: 
    wrap: 512
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)

# Load required packages
library(tidyverse)
library(VariantAnnotation)
library(GenomicRanges)
library(GenomicFeatures)
library(txdbmaker)
library(biomaRt)
library(AnnotationHub)
library(org.Hs.eg.db)
library(rtracklayer)
library(Biostrings)
library(BSgenome)
library(stringr)
library(stringi)
library(gtexr)
library(rlang)
library(glue)
library(readr)

# Create output directory
if (!dir.exists(params$output_dir)) {
  dir.create(params$output_dir, recursive = TRUE)
}

# Source helper functions
source("R/parse_hgvs.R")
source("R/parse_inputs.R")
source("R/utils.R")
source("R/annotations.R")
source("R/breakpoint_annotation.R")
source("R/fusion_prediction.R")
source("R/expression_analysis.R")
source("R/report_generation.R")
source("R/validation.R")

# Initialize logging
log_messages <- list()

log_message <- function(message, level = "INFO") {
  timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  log_entry <- glue::glue("[{timestamp}] [{level}] {message}")
  cat(log_entry, "\n")
  log_messages <<- c(log_messages, list(log_entry))
}

log_message("Fusion Gene Detection Pipeline initialized")
log_message(glue::glue("Input: {params$input_file}"))
log_message(glue::glue("Genome build: {params$genome_build}"))
```

# 1. Project Overview

This Quarto pipeline analyzes structural variants (SVs) from DNA sequencing data to: 1. Parse variants in multiple formats (HGVS, MAVIS, VCF) 2. Identify breakpoints that disrupt genes and transcripts 3. Predict canonical and non-canonical gene fusions 4.
Classify reading frames (in-frame vs out-of-frame) 5. Integrate GTEx expression data to quantify fusion expression patterns 6. Generate comprehensive summary reports and visualizations

**Key Features:** - Supports GRCh38/hg38 genome build - Handles uncertain breakpoint positions - Flags disease genes at breakpoints - Produces publication-ready summary tables and heatmaps

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 2. Data Import and Type Detection

```{r import-input}
log_message("Starting data import phase")

# Detect input type and load data
input_data <- detect_and_parse_input(
  input_file = params$input_file,
  input_format = params$input_format,
  ncbi_contig_mapping_file = params$ncbi_contig_mapping
)

log_message(glue::glue("Input type detected: {input_data$format}"))
log_message(glue::glue("Input records: {nrow(input_data$variants)}"))

# Display first few records
print("First few input records:")
print(head(input_data$variants, 3))
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 3. Chromosome Normalization and Validation

```{r normalize-chromosomes}
log_message("Normalizing chromosome formats and validating coordinates")

# Normalize chromosomes to UCSC style (chr1, chr2, chr3, ..., chrX, chrY, chrM)
normalized_variants <- input_data$variants %>%
  mutate(
    break1_chromosome = normalize_chromosome(break1_chromosome, style = "UCSC"),
    break2_chromosome = normalize_chromosome(break2_chromosome, style = "UCSC")
  )

# Validate chromosomes are valid human chromosomes
valid_chrs <- c(as.character(1:22), "X", "Y", "M") %>%
  gsub("^","chr",.)
invalid_chrs <- unique(c(
  normalized_variants$break1_chromosome[!(normalized_variants$break1_chromosome %in% valid_chrs)],
  normalized_variants$break2_chromosome[!(normalized_variants$break2_chromosome %in% valid_chrs)]
))

if (length(invalid_chrs) > 0) {
  log_message(
    glue::glue("WARNING: Found invalid chromosomes: {paste(invalid_chrs, collapse=', ')}. These variants will be flagged as invalid."),
    level = "WARNING"
  )
}

# Mark invalid variants
normalized_variants <- normalized_variants %>%
  mutate(
    is_valid_chromosome = (break1_chromosome %in% valid_chrs) & (break2_chromosome %in% valid_chrs)
  )
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 4. Assign Unique Variant IDs and Classify Variant Types

```{r assign-ids-and-classify}
log_message("Assigning unique variant IDs and classifying variant types")

# Assign unique 6-digit hex IDs
set.seed(42)  # For reproducibility
variant_ids <- sprintf("%06X", sample(0:16777215, nrow(normalized_variants), replace = FALSE))

variants_processed <- normalized_variants %>%
  mutate(variant_id = variant_ids) %>%
  dplyr::rowwise() %>%
  mutate(
    # Classify variant type based on event_type and breakpoint logic
    variant_type = classify_variant_type(
      event_type = event_type,
      break1_chromosome = break1_chromosome,
      break2_chromosome = break2_chromosome,
      break1_orientation = break1_orientation,
      break2_orientation = break2_orientation
    )
  ) %>%
  dplyr::ungroup() %>%
  relocate(variant_id, variant_type, .before = everything())

log_message(glue::glue("Assigned {nrow(variants_processed)} unique variant IDs"))
print("Variant type distribution:")
print(table(variants_processed$variant_type))
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 5. Handle Uncertain Breakpoint Positions

```{r handle-uncertain-breakpoints}
log_message("Processing uncertain breakpoint positions")

variants_processed <- variants_processed %>%
  dplyr::rowwise() %>%
  mutate(
    # For uncertain breakpoints, use outermost position relative to orientation
    break1_position = handle_uncertain_breakpoint(
      position_start = break1_position_start,
      position_end = break1_position_end,
      orientation = break1_orientation
    ),
    break2_position = handle_uncertain_breakpoint(
      position_start = break2_position_start,
      position_end = break2_position_end,
      orientation = break2_orientation
    ),
    # Flag if breakpoint is uncertain
    break1_uncertain = (break1_position_start != break1_position_end),
    break2_uncertain = (break2_position_start != break2_position_end)
  ) %>%
  dplyr::ungroup()

uncertain_count <- sum(variants_processed$break1_uncertain | variants_processed$break2_uncertain)
if (uncertain_count > 0) {
  log_message(
    glue::glue("Found {uncertain_count} variants with uncertain breakpoint positions. Using outermost coordinates."),
    level = "WARNING"
  )
}
variants_processed
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 6. Load Reference Annotations (GENCODE, Disease Genes, GTEx Metadata)

```{r load-annotations}
log_message("Starting Phase 2: Loading reference annotations")

# Load GENCODE annotations
log_message("Loading GENCODE annotations from AnnotationHub")
gencode_data <- load_gencode_annotations(force_download = FALSE, gencode_release = params$gencode_release)

if (gencode_data$loaded_successfully) {
  log_message(glue::glue("✓ Loaded GENCODE v{gencode_data$gencode_version}"))
  log_message(glue::glue("  Genes: {nrow(gencode_data$genes)}"))
  log_message(glue::glue("  Transcripts: {nrow(gencode_data$transcripts)}"))
  log_message(glue::glue("  Exons: {length(gencode_data$exons)}"))
  log_message(glue::glue("  CDS regions: {length(gencode_data$cds)}"))
} else {
  log_message("✗ Failed to load GENCODE annotations", level = "WARNING")
  log_message(glue::glue("  Error: {gencode_data$error}"), level = "WARNING")
}

# Load disease gene list
log_message("Loading disease gene list")
disease_genes <- load_disease_genes(params$disease_gene_list)

if (nrow(disease_genes) > 0) {
  log_message(glue::glue("✓ Loaded {nrow(disease_genes)} disease genes"))
} else {
  log_message("⚠ No disease genes loaded (file may be missing or empty)", level = "WARNING")
}

# Initialize GTEx metadata
log_message("Initializing GTEx tissue metadata")
gtex_tissues <- gencode_data$gtex_tissues  # Use tissues from annotation load

log_message("✓ Phase 2 complete: All annotations loaded")
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 7. Phase 3: Breakpoint Annotation

```{r phase3-breakpoint-annotation}
log_message("Starting Phase 3: Breakpoint Annotation")

breakpoint_annotations <- annotate_breakpoints(
  variants = variants_processed,
  annotations = gencode_data,
  disease_genes = disease_genes
)

log_message(glue::glue("✓ Annotated {nrow(breakpoint_annotations)} breakpoint-gene pairs"))
log_message(glue::glue("  Genic breakpoints: {sum(breakpoint_annotations$is_gene, na.rm=TRUE)}"))
log_message(glue::glue("  Disease genes flagged: {sum(breakpoint_annotations$is_disease_gene, na.rm=TRUE)}"))

# Summary of breakpoints with exon/intron/cds overlaps
region_summary <- breakpoint_annotations %>%
  dplyr::summarise(
    breakpoints_with_exons = sum(is_exon),
    breakpoints_with_introns = sum(is_intron),
    breakpoints_with_cds = sum(is_cds),
    intergenic_breakpoints = sum(!is_gene)
  )

cat("Breakpoint annotation summary:\n")
print(region_summary)
breakpoint_annotations
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 8. Phase 4: Fusion Prediction

```{r phase4-fusion-prediction}
log_message("Starting Phase 4: Fusion Prediction")

fusions <- predict_fusions(
  breakpoint_annotations = breakpoint_annotations,
  annotations = gencode_data,
  disease_genes = disease_genes
)

log_message(glue::glue("✓ Predicted {nrow(fusions)} fusions"))
log_message(glue::glue("  Canonical: {sum(fusions$is_canonical == TRUE, na.rm=TRUE)}"))
log_message(glue::glue("  Non-canonical: {sum(fusions$is_canonical == FALSE, na.rm=TRUE)}"))
log_message(glue::glue("  In-frame: {sum(fusions$in_frame == TRUE, na.rm=TRUE)}"))
log_message(glue::glue("  Out-of-frame: {sum(fusions$in_frame == FALSE, na.rm=TRUE)}"))

# Display fusion predictions
cat("\nFusion Predictions:\n")
print(fusions %>%
  dplyr::select(fusion_id, variant_type, bp1_gene_symbol, bp2_gene_symbol, 
                is_canonical, in_frame, canonical_fusion_gene_promoter_breakpoint))
fusions
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 9. Phase 5: Expression Analysis

```{r phase5-expression-analysis}
log_message("Starting Phase 5: Expression Analysis")

# Load GTEx expression data
gtex_expression <- load_gtex_expression()

if (!gtex_expression$loaded_successfully) {
  log_message("⚠ GTEx data not available - skipping expression analysis", level = "WARNING")
  log_message("  To use: Download from https://www.gtexportal.org/", level = "WARNING")
  expression_hotspots <- tibble::tibble()
} else {
  log_message(glue::glue("✓ Loaded GTEx v11 expression data"))
  log_message(glue::glue("  {nrow(gtex_expression$expression)} genes × {length(gtex_expression$tissues)} tissues"))
  
  # Identify fusion hotspots
  expression_hotspots <- identify_fusion_hotspots(
    fusion_data = fusions,
    gtex_expression = gtex_expression
  )
  
  n_hotspots <- sum(expression_hotspots$is_hotspot, na.rm = TRUE)
  log_message(glue::glue("✓ Identified {n_hotspots} tissue-level fusion hotspots"))
  
  # Show hotspot summary
  if (n_hotspots > 0) {
    hotspot_summary <- expression_hotspots %>%
      dplyr::filter(is_hotspot) %>%
      dplyr::group_by(fusion_id) %>%
      dplyr::summarise(
        n_hotspot_tissues = dplyr::n(),
        .groups = "drop"
      )
    
    cat("\nHotspots per fusion:\n")
    print(hotspot_summary)
  }
}
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 10. Phase 6: Report Generation

```{r phase6-report-generation}
log_message("Starting Phase 6: Report Generation")

# Generate comprehensive report
report <- generate_fusion_report(
  fusions = fusions,
  expression_hotspots = expression_hotspots,
  breakpoint_annotations = breakpoint_annotations,
  disease_genes = disease_genes
)

log_message("✓ Generated comprehensive fusion report")
log_message(glue::glue("  Fusion summary: {nrow(report$fusion_summary)} entries"))
log_message(glue::glue("  Breakpoint summary: {nrow(report$breakpoint_summary)} entries"))
log_message(glue::glue("  Hotspot summary: {nrow(report$hotspot_summary)} entries"))

# Export tables to TSV
log_message("Exporting report tables to TSV")
export_report_tables(report, output_dir = params$output_dir)

# Generate HTML report
log_message("Generating HTML report")
html_report_file <- file.path(params$output_dir, "fusion_report.html")
generate_html_report(
  report = report,
  output_file = html_report_file,
  title = glue::glue("Fusion Gene Detection Report - {Sys.Date()}")
)

log_message(glue::glue("✓ HTML report written to {html_report_file}"))

# Print report summary
print_report_summary(report)
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 11. Phase 7: Quality Control & Validation

```{r phase7-validation}
log_message("Starting Phase 7: Quality Control & Validation")

if (nrow(fusions) > 0) {
  # Run complete pipeline validation
  validation_result <- validate_complete_pipeline(
    annotations = gencode_data,
    breakpoint_annotations = breakpoint_annotations,
    fusions = fusions,
    expression_hotspots = expression_hotspots,
    report = report
  )
  
  log_message(glue::glue("Validation: {validation_result$total_passed}/{validation_result$total_checks} checks passed"))
  
  if (validation_result$overall_passed) {
    log_message("✓ ALL VALIDATION CHECKS PASSED - Pipeline ready for production use")
  } else {
    log_message("⚠ Some validation checks failed - review details above", level = "WARNING")
  }
  
  # Calculate quality metrics
  quality_metrics <- calculate_quality_metrics(fusions, expression_hotspots, report)
  
  cat("\nData Quality Metrics:\n")
  print(quality_metrics)
} else {
  log_message("⚠ No fusions to validate - skipping Phase 7", level = "WARNING")
}
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 12. Final Diagnostic Summary

```{r final-diagnostic-summary}
log_message("Generating final pipeline summary")

# Input summary
input_summary <- tibble::tibble(
  Category = "Input Data",
  Metric = c("Total variants", "Valid chromosomes", "Variant types"),
  Value = c(
    nrow(variants_processed),
    sum(variants_processed$is_valid_chromosome),
    n_distinct(variants_processed$variant_type)
  )
)

# Annotation summary
annotation_summary <- tibble::tibble(
  Category = "Annotations",
  Metric = c("Genes", "Transcripts", "Exons", "CDS regions"),
  Value = c(
    length(gencode_data$genes_gr),
    nrow(gencode_data$tx_lookup),
    length(gencode_data$exons_gr),
    length(gencode_data$cds_gr)
  )
)

# Analysis summary
analysis_summary <- tibble::tibble(
  Category = "Analysis Results",
  Metric = c("Breakpoints annotated", "Fusions predicted", "Canonical fusions", 
             "In-frame fusions", "Expression hotspots"),
  Value = c(
    nrow(breakpoint_annotations),
    nrow(fusions),
    sum(fusions$canonical == TRUE, na.rm = TRUE),
    sum(fusions$reading_frame == "in-frame", na.rm = TRUE),
    if (nrow(expression_hotspots) > 0) sum(expression_hotspots$is_hotspot, na.rm = TRUE) else 0
  )
)

# Combine and display
all_summary <- dplyr::bind_rows(input_summary, annotation_summary, analysis_summary)

cat("\n========================================\n")
cat("PIPELINE EXECUTION SUMMARY\n")
cat("========================================\n\n")
print(all_summary)

cat("\nVariant Type Distribution:\n")
print(variants_processed %>%
  dplyr::group_by(variant_type) %>%
  dplyr::summarise(count = dplyr::n(), .groups = "drop"))

log_message("✓ Pipeline execution complete")
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 13. Save Final Results

```{r save-final-results}
log_message("Saving final results to {params$output_dir}")

# Ensure output directory exists
if (!dir.exists(params$output_dir)) {
  dir.create(params$output_dir, recursive = TRUE)
}

# Save R objects for downstream use
saveRDS(variants_processed, file.path(params$output_dir, "variants_processed.rds"))
log_message(glue::glue("Saved: variants_processed.rds"))

saveRDS(breakpoint_annotations, file.path(params$output_dir, "breakpoint_annotations.rds"))
log_message(glue::glue("Saved: breakpoint_annotations.rds"))

saveRDS(fusions, file.path(params$output_dir, "fusions.rds"))
log_message(glue::glue("Saved: fusions.rds"))

saveRDS(report, file.path(params$output_dir, "report.rds"))
log_message(glue::glue("Saved: report.rds"))

# Save session information
sink(file.path(params$output_dir, "session_info.txt"))
cat("=== Session Information ===\n\n")
print(sessionInfo())
sink()
log_message(glue::glue("Saved: session_info.txt"))

log_message(glue::glue("Output directory: {params$output_dir}"))
log_message(glue::glue("Generated files:"))
log_message("  - variants_processed.rds")
log_message("  - breakpoint_annotations.rds")
log_message("  - fusions.rds")
log_message("  - report.rds")
log_message("  - tables/ (TSV export files)")
log_message("  - fusion_report.html (interactive HTML report)")
log_message("  - session_info.txt")
```

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 14. Execution Log

```{r execution-log}
cat("\n")
cat("========================================\n")
cat("PIPELINE EXECUTION LOG\n")
cat("========================================\n\n")

for (i in seq_along(log_messages)) {
  cat(log_messages[[i]], "\n")
}

sessionInfo()

cat("\n")
cat("========================================\n")
cat("END OF PIPELINE\n")
cat("========================================\n")
```
